---
title: "R Notes"
params:
  home.dir: /home/fay/code/r_coursera
output:
  html_notebook
---

```{r setup, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(params$home.dir))
```

# Table of Content
* [Search for Documentation]
* [Install/Load Packages]
* [For Loop]
* [Loop Functions]
* [Debugging Tools]
* [Read/Write]
* [Exploratory Data Analysis]
* [Data Wrangling]
* [Plotting]

***

## Search for Documentation
```{r eval=FALSE}
?read.table
help.search('data input')
find('anova')
help(package=lubridate)

args('rnorm')
formals(rnorm)
str(rnorm)
```

[Return to [Table of Content]]

***

## Install/Load Packages
```{r}
a <- available.packages()
head(rownames(a), 3)
```

```{r eval=FALSE}
install.packages('swirl')
install.packages('lubridate')
install.packages('knitr')
```

Install packages from [Bioconductor](http:www.bioconductor.org/install):
```{r eval=FALSE}
source('http://bioconductor.org/biocLite.R')
biocLite(package.name)
```

Install more than one:
```{r eval=FALSE}
biocLite(c('GenomicFeatures', 'AnnotationDbi'))
```

Load a library:
```{r eval=FALSE}
library('swirl')
```

Unload a library:
```{r eval=FALSE}
detach("package:RMySQL", unload=TRUE)
```

Check the version of a package:
```{r}
packageVersion('swirl')
```


Render Rmarkdown:
```{r eval=FALSE}
library('knitr')
```

Run a R script:
```{r eval=FALSE}
source('filename.R')
```

[Return to [Table of Content]]

***

## For Loop
Use `seq_along`:
```{r}
x <- c('a', 'b', 'c', 'd')
for(i in seq_along(x)) {
  print(x[i])
}
```

Use `seq_len`
```{r}
for(i in seq_len(length(x))) {
  print(x[i])
}
```

[Return to [Table of Content]]

***

## Loop Functions
* `apply`
* `lapply`
* `sapply`
* `vapply`
* `mapply`
* `tapply` (related function: `split`)

*Note*: `tapply` splits an atomic object, typically a vector; for a dataframe, use `split` combined with `lapply` or `sapply`.
e.g.
```{r}
library(datasets)
s <- split(airquality, airquality$Month)
lapply(s, function(x) {colMeans(x[, c('Ozone', 'Solar.R', 'Wind')])})
```

Using `tapply` here returns an error:
```{r error=TRUE}
tapply(airquality, airquality$Month, function(x) {colMeans(x[, c('Ozone', 'Solar.R', 'Wind')])})
```
This is because `length(dataframe)` return the number of columns in the dataframe (not the number of rows).

[Return to [Table of Content]]

***

## Debugging Tools
These are interactive tools that are meant to be used in the console:
* `traceback`

```{r error=TRUE}
library(datasets)
tapply(airquality, airquality$Month, function(x) {colMeans(x[, c('Ozone', 'Solar.R', 'Wind')])})
traceback()
```

* `debug`
```{r eval=FALSE}
debug(tapply)
tapply(airquality, airquality$Month, function(x) {colMeans(x[, c('Ozone', 'Solar.R', 'Wind')])})
n  # for the next line
```

* `browser`
* `trace`
* `recover`
```{r eval=FALSE}
options(error = recover)
```

[Return to [Table of Content]]

***

## Read/Write
* [Read/Write R Objects]
* [Download Data]
* [Read/Write Excel Files]
* [Read XML]
* [Read HTML (Webpages)]
* [Read from APIs]
* [Read/Write JSON]
* [Read from MySQL]
* [Read from HDF5]


### Further reading:

- `?connections`
- [This R Data Import Tutorial Is Everything You Need](https://www.r-bloggers.com/this-r-data-import-tutorial-is-everything-you-need/)


### Other database packages:

- `RPostgresql`
- `RODBC`
- `RMongo`

### Read images:

- `jpeg`
- `readbitmap`
- `png`
- `EBImage` (Bioconductor)


### Read GIS data:

- `rdgal`
- `rgeos`
- `raster`


### Read music data:

- `tuneR`
- `seewave`


***

### Read/Write R Objects
#### `dput` & `dget` --- one R object:
```{r eval=FALSE}
y <- data.frame(a = 1, b = 'a')
dput(y, file = 'y.R')
new.y <- dget('y.R')
```

#### `dump` & `source` --- multiple R objects:
```{r eval=FALSE}
x <- 'foo'
y <- data.frame(a = 1, b = 'a')
dump(c('x', 'y'), file = 'data.R')
rm(x, y)
source('data.R')
```

[Return to [Table of Content]]

***

### Download Data
```{r eval=FALSE}
download.file('https://ailuropoda1864.github.io/portfolio/resume_Jingfei_Cai.pdf', destfile = 'resume.pdf')
```

[Return to [Table of Content]]

***

### Read/Write Excel Files
Read Excel files:
```{r eval=FALSE}
library(xlsx)
data <- read.xlsx('filename.xlsx', sheetIndex=1, header=TRUE)
```

Write Excel files:
```{r eval=FALSE}
write.xlsx()
```

[Return to [Table of Content]]

***

### Read XML
```{r}
library(XML)
fileUrl <- 'https://www.w3schools.com/xml/simple.xml'
download.file(fileUrl, destfile = 'example_files/simple.xml')
doc <- xmlTreeParse('example_files/simple.xml', useInternalNodes=TRUE)

rootNode <- xmlRoot(doc)
xmlName(rootNode)
```

#### Directly access parts of the XML document:
```{r}
rootNode[[1]]
rootNode[[1]][[1]]
```

#### Programatically extract parts of the file:
```{r}
xmlSApply(rootNode, xmlValue)
```

#### Extract elements using xpath:
```{r xpath}
xpathSApply(rootNode, '//name', xmlValue)
xpathSApply(rootNode, '//price', xmlValue)
```

[Return to [Table of Content]]

***

### Read HTML (Webpages)

```{r read HTML}
con <- url('https://nationalzoo.si.edu/news/webcam/archive/2238', 'r')
html <- readLines(con)
close(con)

head(html)
```

Parse HTML:
```{r parse HTML}
library(XML)
doc <- htmlTreeParse(html, useInternalNodes = TRUE)
links <- xpathSApply(doc,
                     '//div[@class="views-field views-field-title"]/span[@class="field-content"]/a',
                     xmlGetAttr,
                     name='href')
links
```

Alternatively, use `httr` library:
```{r}
library(httr)
html2 <- GET('https://nationalzoo.si.edu/news/webcam/archive/2238')
content2 <- content(html2, as='text')

parsedHtml <- htmlParse(content2, asText=TRUE)
xpathSApply(parsedHtml,
            '//div[@class="views-field views-field-title"]/span[@class="field-content"]/a',
            xmlGetAttr,
            name='href')
```

**Note:**  `xmlParse` and `htmlParse` are equivalent to the `xmlTreeParse` and `htmlTreeParse` respectively, except they both use a default value for the `useInternalNodes` parameter of TRUE, i.e. they working with and return internal nodes/C-level nodes. These can then be searched using XPath expressions via `xpathApply` and `getNodeSet`.


`GET` in `httr` can be used to access website with passwords:
```{r eval=FALSE}
pg2 = GET(url.string, authenticate('user', 'passwd'))
```

Use handles to preserve preserves settings and cookies across multiple requests (e.g. authentication):
```{r eval=FALSE}
google <- handle('http://google.com')
pg1 <- GET(handle = google, path='/')
pg2 <- GET(handle = google, path='search')
```

[Return to [Table of Content]]

***

### Read from APIs
```{r eval=FALSE}
library(httr)
myapp <- oauth_app('twitter', key='consumerKey', secret='consumerSecret')
sig <- sign_oauth1.0(myapp, token='token', token_secret='tokenSecret')
homeTL <- GET('https://api.twitter.com/1.1/statuses/home_timeline.json', sig)
json1 <- content(homeTL)

# use the jsonlite library to reformat the json object so that it's easier to read
json2 <- jsonlite::fromJSON(toJSON(json1))
```

[Return to [Table of Content]]

***

### Read/Write JSON
#### Read JSON:
```{r}
library(jsonlite)
jsonData <- fromJSON('https://api.github.com/users/Ailuropoda1864/repos')
names(jsonData)
```

#### Write data frames to JSON (result hidden):
```{r write to JSON, results='hide'}
library(datasets)
myjson <- toJSON(iris, pretty=TRUE)  # pretty print
cat(myjson)
```

#### Convert back to JSON:
```{r}
iris2 <- fromJSON(myjson)
head(iris2)
```

[Return to [Table of Content]]

***

### Read from MySQL
Functions covered:

* `dbConnect`
* `dbDisconnect`
* `dbListTables`
* `dbReadTable`
* `dbListFields`
* `dbGetQuery`
* `dbSendQuery`
* `dbFetch`
* `dbClearResult`

Further reading:

- [MySQL and R](https://www.r-bloggers.com/mysql-and-r/)

Connect to and query a MySQL server:
```{r}
library(RMySQL)
ucscDb <- dbConnect(MySQL(), user='genome', host='genome-mysql.soe.ucsc.edu')
result <- dbGetQuery(ucscDb, 'show databases;')
dbDisconnect(ucscDb)

head(result)
```

Connect to a specific database:
```{r}
hg19 <- dbConnect(MySQL(), user='genome', host='genome-mysql.soe.ucsc.edu',
                  db='hg19')
```

List all tables:
```{r}
allTables <- dbListTables(hg19)
length(allTables)
allTables[1:5]
```

Get dimensions of a specific table:
```{r}
dbListFields(hg19, 'affyU133Plus2')
```

Make a query:
```{r}
dbGetQuery(hg19, 'SELECT count(*) FROM affyU133Plus2;')
```

Read from a specific table:
```{r warning=FALSE}
affyData <- dbReadTable(hg19, 'affyU133Plus2')
head(affyData)
```

Select a specific subset:
```{r warning=FALSE}
query.string <-
'SELECT * from affyU133Plus2
WHERE misMatches BETWEEN 1 AND 3'

affyMis <- dbGetQuery(hg19, query.string, n=10)
dim(affyMis)
```

**Note:**
`dbGetQuery()` is equivalent to `dbSendQuery()`, then `dbFetch()`, ensuring that the result is always free-d by `dbClearResult()`.
```{r eval=FALSE}
query.string <-
'SELECT * from affyU133Plus2
WHERE misMatches BETWEEN 1 AND 3'

query <- dbSendQuery(hg19, query.string)
affyMis <- fetch(query, n=10)
dbClearResult(query)

dim(affyMis)
```

Close connection when done:
```{r}
dbDisconnect(hg19)
```

[Return to [Table of Content]]

***

### Read from HDF5
Install R HDF 5 package:
```{r eval=FALSE}
source('http://bioconductor.org/biocLite.R')
biocLite('rhdf5')
```

Create a HDF5 file:
```{r}
library(rhdf5)
h5.file <- 'example_files/example.h5'
created <- h5createFile(h5.file)
created
```

Create groups:
```{r}
created <- h5createGroup(h5.file, 'foo')
created <- h5createGroup(h5.file, 'baa')
created <- h5createGroup(h5.file, 'foo/foobaa')
h5ls(h5.file)
```

Write to groups:
```{r}
A <- matrix(1:10, nrow = 5)
h5write(A, h5.file, 'foo/A')

B <- array(seq(0.1, 2.0, by=0.1), dim=c(5,2,2))
attr(B, 'scale') <- 'liter'
h5write(B, h5.file, 'foo/foobaa/B')

h5ls(h5.file)
```

Read from a HDF5 file:
```{r}
readA <- h5read(h5.file, 'foo/A')
readB <- h5read(h5.file, 'foo/foobaa/B')

readA
```

Read/write in chunks:
```{r}
h5write(c(12, 13, 14), h5.file, 'foo/A',
        index=list(1:3, 1))  # write to the first 3 rows in the first column
h5read(h5.file, 'foo/A')

h5read(h5.file, 'foo/A', index=list(1:3, 2))
```


[Return to [Table of Content]]

***

## Exploratory Data Analysis

#### What the actual data looks like
```{r}
data("UCBAdmissions")
df <- as.data.frame(UCBAdmissions)

head(df)
tail(df, 3)
```

#### Numbers of rows/columns, names of rows/columns, the class of each column, any missing values, any duplicates
```{r}
str(df)  # info() in Python
rownames(df)  # df.index in Python
names(df)  # df.columns in Python

# missing values
sum(is.na(df$Admit))
colSums(is.na(df))

# duplicates
sum(duplicated(df))
```

#### Distributions
- Quantitative variables: quantiles
- Qualatative variables: number of levels; count for each level
```{r}
summary(df)  # describe() in Python

quantile(df$Freq, na.rm=TRUE, probs=c(0, 0.25, 0.5, 0.75, 1))

table(df$Admit, useNA='ifany')  # value_counts in Python
table(df$Gender, df$Dept)
```

#### Crosstab
```{r}
xtabs(Freq ~ Gender + Admit, data=df)
```

[Return to [Table of Content]]

***

## Data Wrangling
* [Handle Missing Values]
* [Factor]
* [Text]
* [RegEx]
* [Datetime]
* [`data.table`]
* [`plyr`]
* [`dplyr`]
* [`tidyr`]
* [Reshape Data]
* [Merge Data]

### Handle Missing Values
```{r}
x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x)
x[!bad]
```

```{r}
y <- c('a', 'b', 'd', NA, NA, 'f')
good <- complete.cases(x, y)
x[good]
```

`complete.cases` can also be used with dataframes:
`complete.cases(dataframe_object)`

When subsetting a vector/data frame containing `NA` using conditional, use `which`:
```{r}
x[which(x>2)]
# compare with
x[x>2]
```

[Return to [Table of Content]]

***

### Factor
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"), levels = c("yes", "no"))
x
```

```{r}
unclass(x)
```

```{r}
relevel(x, ref='no')
```

#### Generate factor levels
```{r}
f <- gl(2, 5)
f
```

#### Compute factor interactions
```{r}
x <- rnorm(10)
f1 <- gl(5, 2)
f2 <- gl(2, 5)
interaction(f1, f2)
```

#### Split
When more than one factors are put in a list, `split` will call `interaction` automatically:
```{r}
split(x, list(f1, f2))
```

Drop empty levels:
```{r}
split(x, list(f1, f2), drop = TRUE)
```

#### Create binary variables:
```{r eval=FALSE}
ifelse(test, yes, no)
```

#### Create categorical variables
```{r eval=FALSE}
cut(x, breaks)

# Alternatively
library(Hmisc)
cut2(x, g=4)
```

### Text

- `tolower` / `toupper`
- `strsplit`
- `sub` / `gsub`  
can be RegEx 
- `grep` (return indices or values) / `grepl` (return T/F)  
can be RegEx
- `nchar`
- `substr`
- `paste` / `paste0`
- `stringr::str_trim`

[Return to [Table of Content]]

***

### RegEx

metacharacters:

- `^`  
beginning
- `$`  
end
- `.`  
any character
- `|`  
alternative patterns

escape:  
`\\`

[Return to [Table of Content]]

***

### Datetime
```{r eval=FALSE}
library(lubridate)
today()
t1 <- now(tzone = "America/New_York")

update(now(), hours=14, minutes=3)
now() + hours(1) + minutes(2)

with_tz(now(), tzone='Asia/Hong_Kong')

time.between <- interval(t1, now())
as.period(time.between)
```

List of valid time zones for use with `lubridate`:  
http://en.wikipedia.org/wiki/List_of_tz_database_time_zones


[Return to [Table of Content]]

***

### `data.table`
```{r load data.table}
library(data.table)
```

```{r}
DT <- data.table(x=rnorm(9), y=rep(c('a', 'b', 'c'), each=3), z=rnorm(9))
head(DT, 3)
```

See all the data tables in memory:
```{r}
tables()
```

Subset rows:
```{r}
DT[2,]
DT[DT$y=='a',]
DT[c(2,3)]
```

Columns expressions:
```{r}
DT[, c(2,3)]
DT[, list(mean(x), sum(z))]
DT[, table(y)]

# add new columns
DT[, w:= z^2]

# multiple operations
DT[, m:= {tmp <- (x+z); log2(tmp+5)}]

# plyr-like operations
DT[, a:= x>0]
DT[, b:= mean(x+w), by=a]

# special variables
# .N is like count
DT[, .N, by=a]
```

Keys
```{r}
DT <- data.table(x=rep(c('a', 'b', 'c'), each=100), y=rnorm(300))
setkey(DT, x)
DT['a']
```

Use keys to facilitate joins:
```{r}
DT1 <- data.table(x=c('a', 'a', 'b', 'dt1'), y=1:4)
DT2 <- data.table(x=c('a', 'b', 'dt2'), y=5:7)
setkey(DT1, x)
setkey(DT2, x)
merge(DT1, DT2)
```

Read in file:
```{r eval=FALSE}
fread(filename)
```

[Return to [Table of Content]]

***

### `plyr`
```{r load plyr}
library(plyr)
```

#### Order a dataframe by certain variables
```{r eval=FALSE}
arrange(dataframe, var1)
# use base R:
dataframe[order(dataframe$var1), ]

# in descending order:
arrange(dataframe, desc(var1))
```

#### Copy a data frame and add new variables
```{r eval=FALSE}
df2 <- mutate(df1, newCol=cut2(oldCol, g=4))
```

#### Split-apply-combine
```{r}
str(InsectSprays, 3)
ddply(InsectSprays, .(spray), summarize, sum=sum(count))
```

Or, maintain the dimension so that a new variable can be created:
```{r}
spraySums <- ddply(InsectSprays, .(spray), summarize, sum=ave(count, FUN=sum))
dim(spraySums)
head(spraySums)
```

#### Merge data
Compared with `merge` in base R, it is faster, but less full featured (defaults to left join)
```{r eval=FALSE}
join(x, y, by = NULL, type = "left", match = "all")
```

Merge multiple data frames (which is more difficult with `merge`)
```{r}
df1 <- data.frame(id=sample(1:10), x=rnorm(10))
df2 <- data.frame(id=sample(1:10), y=rnorm(10))
df3 <- data.frame(id=sample(1:10), z=rnorm(10))
dfList = list(df1, df2, df3)
join_all(dfList)
```

[Return to [Table of Content]]

***

### `dplyr`
Convert a data frame to a "data frame tbl" object (for easier viewing):
```{r eval=FALSE}
tbl_df(dataframe)
```

- `select`
subsetting columns
- `filter`
subsetting rows
- `arrange`
- `rename`  
`rename(df, newName = oldName)`
- `mutate`
- `group_by`
- `summarize`
    - `n()`  
    total number of rows
    - `n_distinct(var)`  
    total number of unique rows
- `%>%`

[Return to [Table of Content]]

***

### `tidyr`
#### When column headers are values, not variable names
`gather`

#### When multiple variables are stored in one column
`separate`  
default to split on non-alphanumeric values (the `sep` argument)

#### When variables are stored in both rows and columns
`spread`

#### When multiple observational units are stored in the same table
`select` followed by `unique`

#### When a single observational unit is stored in multiple tables
`bind_rows`

[Return to [Table of Content]]

***

### Reshape Data
```{r load reshape2, eval=FALSE}
library(reshape2)
```

#### Melt data frames
```{r eval=FALSE}
melt(df, id, measure.vars)
```

#### Cast data frames
```{r eval=FALSE}
dcast(data, formula, fun.aggregate)
```
See also: `acast`

[Return to [Table of Content]]

***

### Merge Data
`merge`  
`intersect`


[Return to [Table of Content]]

***

## Plotting
* [Base]
* [lattice]
* [ggplot2]
* [Graphics Devices]

### Base
Use `example(points)` to see examples (press `Esc` to exit).

#### Base graphics parameters
Use `?par` to see the complete list. Also: `names(par())`
  
To see the default values for global graphics paremeters:
```{r}
par('lty')
```

- `pch`  
The plottin symbol (default is open circle).
  
- `lty`  
The line type. Line types can either be specified as an integer (0=blank, 1=solid (default), 2=dashed, 3=dotted, 4=dotdash, 5=longdash, 6=twodash) or as one of the character strings "blank", "solid", "dashed", "dotted", "dotdash", "longdash", or "twodash", where "blank" uses ‘invisible lines’ (i.e., does not draw them).
  
- `col`  
The plotting color. The `colors()` function provides a vector of colors by name.

- `las`  
The orientation of the axis labels on the plot.

- `mfrow`  
Number of plots per row, column (plots are filled row-wise).

- `mfcol`  
Number of plots per row, column (plots are filled column-wise).
  
The par() function is used to specify global graphics parameters that affect all plots in an R session. (Use `dev.off` or `plot.new` to reset to the defaults.)

#### Scatterplot
```{r}
library('datasets')

head(cars)
plot(cars, main='Speeds of cars by distance')

head(Puromycin)
plot(Puromycin$conc, Puromycin$rate, col=Puromycin$state)
# add regression line
model <- lm(rate ~ conc, Puromycin)
abline(model, col='blue')


head(airquality)
with(airquality,
     plot(Wind, Ozone, 
          main='Ozone and Wind in New York City',
          type='n'))  # 'n' for no plotting
with(subset(airquality, Month==5), points(Wind, Ozone, col='blue'))
with(subset(airquality, Month!=5), points(Wind, Ozone, col='red'))
legend('topright', pch=1, 
       col=c('blue', 'red'), 
       legend=c('May', 'Other Months'))
```

Another example of customerizing `legend`:
```{r eval=FALSE}
legend('topright', 
       legend=paste0('Sub_metering_', 1:3), 
       col=c("black","red","blue"), 
       lty=c(1,1,1),  # the symbols are lines
       bty='n')  # no border around legend

```

Other functions that add to a plot (but do not create a plot):  

- `lines`
- `points`
- `text`
- `title`
- `mtext`  
add arbitrary text to the margines (inner or outer) of the plot
- `axis`

#### Boxplot
```{r}
boxplot(cars)

# use formula
boxplot(Temp ~ Month, data = airquality, xlab = 'Month', ylab = 'Ozone (ppb)')
# add horizontal line
abline(h=mean(airquality$Temp, na.rm = TRUE), lwd=2, col='red', lty=2)  
```

#### Histogram
```{r}
hist(cars$speed, col='green', breaks=10)
rug(cars$speed)
# add vertical line
abline(v=mean(cars$speed, na.rm=TRUE), lwd=4)
```

#### Barplot
```{r}
barplot(table(Puromycin$state))
```

#### Create subplots:
```{r eval=FALSE}
par(mfcol=c(1,2))
```

#### Use `with` to simplify code:
```{r echo=-1}
par(mfcol=c(1,2))
with(cars, {
  boxplot(dist, xlab='distance')
  title('Stopping distance (ft)')
  boxplot(speed)
  title('Speed (mph)')
}
     )
```

[Return to [Table of Content]]

***

### lattice
```{r}
library(lattice)
state <- data.frame(state.x77, region=state.region)
head(state)
xyplot(Life.Exp ~ Income | region, data=state, layout=c(4,1))
```

Also:
`bwplot`

[Return to [Table of Content]]

***

### ggplot2
```{r}
library(ggplot2)
qplot(conc, rate, data=Puromycin)
```

[Return to [Table of Content]]

***

### Graphics Devices
Use `?Devices` to see the graphics devices available on current system.

```{r eval=FALSE}
# open file dvice; create pdf file in wd
pdf(file='myplot.pdf')
# create plot
plot(x, y)
# add annotation
title(main='Title')
# close the pdf file device
dev.off()
# now the pdf file can be viewed on computer
```

Find the currently active graphics device:
```{r eval=FALSE}
dev.cur()
```

Change the active graphics device:
```{r eval=FALSE}
dev.set(<integer>)
```

Copy plots:
```{r eval=FALSE}
# copy a plot from one device to another
dev.copy(png, file='myplot.png')
# close the png device
dev.off()

# copy a plot to a pdf file
dev.copy2pdf()
```



[Return to [Table of Content]]

***